---
title: "ST 558 Project One"
author: "Robert Berini & Robert Ferrand"
format: html
editor: visual
---
## Project Overview

some placeholder text

## Required Packages

Load required packages for project
```{r}
library(tidyverse)
library(httr)
library(jsonlite)
library(tidycensus)
```


## Data Processing

some placeholder text

```{r}

# function for creating data should assign a class of census
# class(your_tibble) <- c("census", class(your_tibble))
```


## Obtaining Person Level Records

some placeholder text

```{r}

```


## Writing a Generic Function for Summarizing

some placeholder text

### Numeric Summary
write a function that produces means and standard deviations for our numeric variable(s) and counts for our categorical variable(s).

```{r}

```

Test out this function by running summary(_your_census_tibble_) on something you’ve returned from your census API function.

```{r}

```

### Graphical Summary

Similarly, let’s create a generic plot() function for a census class tibble. Require the user to specify one categorical variable and one numeric variable for plotting purposes.

```{r}

#ggplot(your_census_tibble, aes(x = get(cat_var), y = get(num_var), weight = PWGTP)) + geom_boxplot()
```

